Functory: A Distributed Computing Library \\ for Objective Caml \thanks{This research was partly supported by the French national project U3CAT (\emph{Unification of Critical C Code Analysis Techniques}, ANR-08-SEGI-021).}

### 2011-09-16 doc/tfp_2011_final/main.tex
started revision KK

% \documentclass[tfpsymp,pagenumbers]{tfp07symp}
\documentclass{llncs}

@@

\newcommand{\Ocaml}{OCaml}
\newcommand{\functory}{\textsf{Functory}}
\newcommand{\JoCaml}{Jo{\&\!}Caml}
\newcommand{\unix}{\textsc{Unix}}

@@

\paragraph{Related Work.}
In order to compare and better distinguish
\of{Functory} from others work with related goals and motivations, we can
broadly classify the related work in this domain into:
\begin{enumerate}
\item \textit{Distributed Functional Languages (DFLs)} --- functional
  languages that provide built-in primitives for
  distribution. Examples include ML5, \JoCaml\, Glasgow Distributed
  Haskell, Erlang, etc.
\item \textit{Libraries for existing functional languages} --- that
  could be readily used in order to avoid implementing details like
  task distribution, fault-tolerance, socket programming, etc.
\end{enumerate}
\functory\ belongs to the second category.
For reasons of completeness, though, we first describe some existing DFLs
related to functional programming.

\JoCaml\ is one of the DFLs which provides communication primitives
(like channels) for facilitating transmission of
computations. However, it does not provide ready-made language
features for fault-tolerance, which is indispensable in a distributed
setting. The user has to include code for fault-tolerance, as already
demonstrated in some \JoCaml\
library~\cite{mandel2008}. ML5~\cite{ML5}, a variant of ML, is a
programming language for distributed computing, specialized for web
programming. It provides primitives for transferring control between
the client and the server, as well as low-level primitives for
marshalling the data. As in the case before, ML5 is a programming
language that offers primitives for code mobility, and the code for
distribution of computation and fault-tolerance has to be included by
the user.  ML5 implements type-safe marshalling and
\functory\ does not, though an existing type-safe marshalling library
could be used with \functory. 
% Basically, the task will be transmitted
% along with the corresponding type and on unmarshalling, it could be
% matched for the correct type.
Glasgow Distributed Haskell (GdH)~\cite{GdH} is a pure distributed
functional language that is built on top of Glasgow Haskell and
provides features for distributed computing. It is an extension of
both Glasgow Parallel Haskell, that supports only one process and
multiple threads and Concurrent Haskell that supports multiple
processes. It also offers features for fault-tolerance - error
detection and error recovery primitives in the
language.

@@

The iTask system~\cite{iTask} is a library for the functional language
`Clean' targeted at distributed workflow management. The library
provides a set of combinators (some of which perform map/fold
operations) that facilitate applications running in different nodes
of a distributed system to communicate, exchange information and
coordinate their computations.

\begin{comment}
Closest to the approach in this paper is Yohann Padioleau's MapReduce
implementation in \Ocaml~\cite{poor-man-mapreduce}.  It is built on
top of OCamlMPI~\cite{ocamlMPI}, while our approach uses a homemade
protocol for message passing.  Currently, we have less flexibility
w.r.t. deployment of the user program than OCamlMPI; on the other
hand, we provide a 
more generic API together with fault-tolerance.  There are other
distributed computing libraries on top of which one could implement
the library discussed in this paper. \JoCaml~\cite{jocaml} is one of
them. However, \JoCaml\ does not provide fault-tolerance, which is
indispensable in a distributed setting. The user has to include code
for fault-tolerance, as already demonstrated in some \JoCaml\
experiments~\cite{mandel2008}.

@@

In most cases, the easiest way to parallelize an execution it to make
use of operations over lists, where processing of the
list elements are done in parallel. 
To facilitate such a processing,
our library provides most commonly used list operations, all 
implemented using our generic \of{compute} function.

@@

In the case of \of{map_remote_fold}, only one \of{fold} operation can
be performed at a time (possibly in parallel with \of{f}
operations), as obvious from~(\ref{eq:map-fold}). However, there are
cases where several \of{fold} operations can be performed in parallel,
as early as intermediate results of \of{fold} operations are available.
This is the case when \of{fold} is an associative operation (which
implies that types \of{'b} and \of{'c} are the same).
Whenever
\of{fold} is also commutative, we can perform even more \of{fold}
operations in parallel. Thus our API provides two functions
\of{map_fold_a} and \of{map_fold_ac} for these two particular cases,
with types
\begin{ocaml}
  val map_fold_ac, map_fold_a: 
    f:('a -> 'b) -> fold:('b -> 'b -> 'b) -> 'b -> 'a list -> 'b 
\end{ocaml}
It is rather straightforward to derive these five functions from the
generic \of{compute} function; readers interested in details could
refer to the source code.

@@

  As mentioned in Section~\ref{sec:API}, the \of{Network} module
  actually provides three different implementations, according to
  three different execution scenarios. There are provided in three
  sub-modules, as described below.

@@

The purpose of the following experiments is to compare the various
deployments, namely sequential, cores and network.
For this comparison to be fair, all
computations are performed on the same machine, a 8 core Intel Xeon
3.2 GHz running Debian Linux. The sequential implementation uses a
single core. The multi-core implementation uses up to 8 cores of
the machine. The network implementation uses 8 workers running locally and
a master running on a remote machine over a LAN 
(which incurs communication cost).

@@

The first example is the classical $N$-queens problem, where we
compute the total number of ways to place $N$ queens on a $N\times N$
chessboard in such a way no two queens attack each other.
We use a standard backtracking algorithm for this problem, which
places the queens one by one starting from the first row.
Distributing the computation is thus quite easy: we consider all
possible ways to place queens on the first $D$ rows and then perform
the subsequent search in parallel. Choosing $D=1$ will result in
exactly $N$ tasks; choosing $D=2$ will result in $N^2-3N+2$ tasks; 
greater values for $D$ would result in two many tasks.

@@

We compare the performances of two different implementations. In the
first one, called \of{mm1}, each task consists of the computation of a single
coefficient of the resultant matrix. 
In the second one, called \of{mm2}, each task consists of the
computation of a whole row of the resultant matrix.
As a consequence, the total number of tasks in
$10,000$ for \of{mm1} and only $100$ for \of{mm2}.
On the contrary, each task result for \of{mm1} is a single integer,
while for \of{mm2} it is a row of 100 integers.
%
The experimental results (in seconds) are tabulated below.
\begin{center}
  \begin{tabular}{|r|r|r|}
    \hline
    & \of{mm1}       & \of{mm2}  \\
    & (10,000 tasks) & (100 tasks) \\
    \hline\hline
\of{Sequential} & 20.3 &  20.2 \\
\hline
 \of{Cores} 
 (2 cores)     &   22.7 (0.89$\times$) &  11.3 (1.79$\times$) \\
 (4 cores)     &   12.3 (1.65$\times$) &   6.1 (3.31$\times$) \\
 (6 cores)     &    8.6 (2.36$\times$) &   4.3 (4.70$\times$) \\
 (8 cores)     &    8.0 (2.54$\times$) &   3.5 (\underline{5.77}$\times$) \\
 \hline
  \end{tabular}
\end{center}
The difference in the number of tasks explains the differences in the
speedup ratios above.
We do not include results for the network configuration, as they do
not achieve any benefit with respect to the sequential
implementation. The reason is that the communication cost dominates the
computation cost in such a way that the total execution time is
always greater than 30 seconds. Indeed, irrespective of the
implementation (\of{mm1} or \of{mm2}), the total size of the
transmitted data is $10^6$ integers, which in our case amounts to
billions of bytes.

@@

Drawing the Mandelbrot set is another classical example that could be
distributed easily, since the color of each point can be computed
independently of the others. 
This benchmark consists in drawing the fragment of the Mandelbrot set
with lower left corner $(-1.1, 0.2)$ and upper right corner $(-0.8,
0.4)$, as a $9,000\times6,000$ image. 
%
If the total number of tasks $\of{t} \ge 1$ is given as a
parameter,  it is immediate to split the image into \of{t} sub-images,
each of which is computed in parallel with and independently of the
others. In our case, the image is split into horizontal slices.
Each task is thus four floating-point numbers denoting the region
coordinates, together with two integers denoting the dimensions of the
sub-image to be drawn. The result of the task is a matrix of pixels,
of size $54,000,000 / \of{t}$. 
For instance, using $\of{t}=20$ tasks will
result in 20 sub-images of size $10.3$ Mb each,
assuming each pixel is encoded in four bytes.

@@

% TODO compute ratios for Network
\begin{center}
  \begin{tabular}{|r|r|r|r|r|}
    \hline
    \#cores  &\#tasks & \of{Cores} & \of{Network} \\
    \hline\hline
    2       & 10 & 15.8       (1.86$\times$) &  20.3  (1.45$\times$)      \\
            & 30 & 15.7       (\underline{1.87}$\times$) &  18.7 (1.57$\times$)       \\
            & 100 & 16.1      (1.83$\times$) &  19.8   (1.48$\times$)    \\
            & 1000 & 19.6     (1.50$\times$) &  38.6    (0.76$\times$)  \\
    \hline
    4       & 10 & 9.50       (3.09$\times$)  &  14.4     (2.04$\times$)  \\
            & 30 & 8.26       (\underline{3.56}$\times$)  &  11.4  (2.58$\times$) \\
            & 100 & 8.37      (3.51$\times$)  &  11.4  (2.58$\times$) \\
            & 1000 & 10.6     (2.77$\times$)  &  20.5   (1.43$\times$) \\
    \hline
    8       & 10 & 9.40       (3.13$\times$)  &  12.6    (2.33$\times$)  \\
            & 30 & 4.24       (\underline{6.93}$\times$)  &   7.6  (3.87$\times$)    \\
            & 100 & 4.38      (6.71$\times$)  &   7.5    (3.92$\times$)  \\
            & 1000 & 6.86     (4.29$\times$)  &  11.3    (2.60$\times$)  \\
    \hline
  \end{tabular}
\end{center}
The best timings are achieved for the \of{Cores} configuration, where
communications happen within the same machine and are thus cheaper.
There are two significant differences with respect to the n-queens
benchmark.  On one hand, the number of tasks can be controlled more
easily than in the case of n-queens. We experimentally figured out the
optimal number of tasks to be 30. One the other hand, each computation
result is an image, rather than just an integer as in the case of
n-queens. Consequently, communication costs are much greater. 
In this particular experiment, the total size of the results
transmitted is more than 200 Mb.

>>>>

% \documentclass[tfpsymp,pagenumbers]{tfp07symp}

\documentclass{llncs}

@@

\newcommand{\Ocaml}{OCaml}
\newcommand{\functory}{\textsf{Functory}}
%\newcommand{\JoCaml}{Jo{\&\!}Caml}
\newcommand{\unix}{\textsc{Unix}}

@@

\paragraph{Related Work.}
In order to compare and better distinguish
\of{Functory} from others work with related goals and motivations, we can
broadly classify the related work in this domain into:
\begin{enumerate}
\item \textit{Distributed Functional Languages (DFLs)} --- functional
  languages that provide built-in primitives for
  distribution. Examples include ML5, JoCaml, Glasgow Distributed
  Haskell, Erlang, etc.
\item \textit{Libraries for existing functional languages} --- that
  could be readily used in order to avoid implementing details like
  task distribution, fault-tolerance, socket programming, etc.
\end{enumerate}
\functory\ belongs to the second category.
For reasons of completeness, though, we first describe some existing DFLs
related to functional programming.

JoCaml is one of the DFLs which provides communication primitives
(like channels) for facilitating transmission of
computations. However, it does not provide ready-made language
features for fault-tolerance, which is indispensable in a distributed
setting. The user has to include code for fault-tolerance, as already
demonstrated in some JoCaml
library~\cite{mandel2008}. ML5~\cite{ML5}, a variant of ML, is a
programming language for distributed computing, specialized for web
programming. It provides primitives for transferring control between
the client and the server, as well as low-level primitives for
marshalling the data. As in the case before, ML5 is a programming
language that offers primitives for code mobility, and the code for
distribution of computation and fault-tolerance has to be included by
the user.  ML5 implements type-safe marshalling and
\functory\ does not, though an existing type-safe marshalling library
could be used with \functory. 
% Basically, the task will be transmitted
% along with the corresponding type and on unmarshalling, it could be
% matched for the correct type.
Glasgow Distributed Haskell (GdH)~\cite{GdH} is a pure distributed
functional language that is built on top of Glasgow Haskell and
provides features for distributed computing. It is an extension of
both Glasgow Parallel Haskell, that supports only one process and
multiple threads and Concurrent Haskell that supports multiple
processes. It also offers features for fault-tolerance - error
detection and error recovery primitives in the
language.

@@

The iTask system~\cite{iTask} is a library for the functional language
`Clean' targeted at distributed workflow management. The library
provides a set of combinators (some of which perform map/fold
operations) that facilitate applications running in different nodes
of a distributed system to communicate, exchange information and
coordinate their computations in a type-safe manner.

\begin{comment}
Closest to the approach in this paper is Yohann Padioleau's MapReduce
implementation in \Ocaml~\cite{poor-man-mapreduce}.  It is built on
top of OCamlMPI~\cite{ocamlMPI}, while our approach uses a homemade
protocol for message passing.  Currently, we have less flexibility
w.r.t. deployment of the user program than OCamlMPI; on the other
hand, we provide a 
more generic API together with fault-tolerance.  There are other
distributed computing libraries on top of which one could implement
the library discussed in this paper. JoCaml~\cite{jocaml} is one of
them. However, JoCaml does not provide fault-tolerance, which is
indispensable in a distributed setting. The user has to include code
for fault-tolerance, as already demonstrated in some JoCaml
experiments~\cite{mandel2008}.

@@

In most cases, the easiest way to parallelize an execution is to make
use of operations over lists, where processing of the
list elements are done in parallel. 
To facilitate such a processing,
our library provides most commonly used list operations, all 
implemented using our generic \of{compute} function.

@@

In the case of \of{map_remote_fold}, only one \of{fold} operation can
be performed at a time (possibly in parallel with \of{f}
operations), as obvious from~(\ref{eq:map-fold}). However, there are
cases where several \of{fold} operations can be performed in parallel,
as early as intermediate results of \of{fold} operations are available.
This is the case when \of{fold} is an associative operation (which
implies that types \of{'b} and \of{'c} are the same).
Whenever
\of{fold} is also commutative, we can perform even more \of{fold}
operations in parallel. Thus our API provides two functions
\of{map_fold_a} and \of{map_fold_ac} for these two particular cases,
with types
\begin{ocaml}
  val map_fold_ac, map_fold_a: 
    f:('a -> 'b) -> fold:('b -> 'b -> 'b) -> 'b -> 'a list -> 'b 
\end{ocaml}
It is rather straightforward to derive these five functions from the
generic \of{compute} function; we invite readers interested in details to
refer to the source code.

@@

  As mentioned in Section~\ref{sec:API}, the \of{Network} module
  actually provides three different implementations as sub-modules, according to
  three different execution scenarios, the details of which are
  presented below:

@@

The purpose of the following experiments is to compare the various
deployments, namely sequential, cores and network.
For this comparison to be fair, all
computations are performed on the same machine, an 8 core Intel Xeon
3.2 GHz running Debian Linux. The sequential implementation uses a
single core. The multi-core implementation uses up to 8 cores of
the machine. The network implementation uses 8 workers running locally and
a master running on a remote machine over a LAN 
(which incurs communication cost).

@@

The first example is the classical $N$-queens problem, where we
compute the total number of ways to place $N$ queens on a $N\times N$
chessboard in such a way no two queens attack each other.
We use a standard backtracking algorithm for this problem, which
places the queens one by one starting from the first row.
Distributing the computation is thus quite easy: we consider all
possible ways to place queens on the first $D$ rows and then perform
the subsequent search in parallel. Choosing $D=1$ will result in
exactly $N$ tasks; choosing $D=2$ will result in $N^2-3N+2$ tasks; 
greater values for $D$ would result in too many tasks.

@@

We compare the performances of two different implementations. In the
first one, called \of{mm1}, each task consists of the computation of a single
coefficient of the resultant matrix. 
In the second one, called \of{mm2}, each task consists of the
computation of a whole row of the resultant matrix.
As a consequence, the total number of tasks is
$10,000$ for \of{mm1} and only $100$ for \of{mm2}.
On the contrary, each task result for \of{mm1} is a single integer,
while for \of{mm2} it is a row of 100 integers.
%
The experimental results (in seconds) are tabulated below.
\begin{center}
  \begin{tabular}{|r|r|r|}
    \hline
    & \of{mm1}       & \of{mm2}  \\
    & (10,000 tasks) & (100 tasks) \\
    \hline\hline
\of{Sequential} & 20.3 &  20.2 \\
\hline
 \of{Cores} 
 (2 cores)     &   22.7 (0.89$\times$) &  11.3 (1.79$\times$) \\
 (4 cores)     &   12.3 (1.65$\times$) &   6.1 (3.31$\times$) \\
 (6 cores)     &    8.6 (2.36$\times$) &   4.3 (4.70$\times$) \\
 (8 cores)     &    8.0 (2.54$\times$) &   3.5 (\underline{5.77}$\times$) \\
 \hline
  \end{tabular}
\end{center}
The difference in the number of tasks explains the differences in the
speedup ratios above.
We do not include results for the network configuration, as they do
not achieve any benefit with respect to the sequential
implementation. The reason is that the communication cost dominates the
computation cost in such a way that the total execution time is
always greater than 30 seconds. Indeed, irrespective of the
implementation (\of{mm1} or \of{mm2}), the total size of the
transmitted data is $10^6$ integers, which in our case amounts to
billions of bytes.

@@

Drawing the Mandelbrot set is another classical example that could be
distributed easily, since the color of each point can be computed
independently of the others. 
This benchmark consists in drawing the fragment of the Mandelbrot set
with lower left corner $(-1.1, 0.2)$ and upper right corner $(-0.8,
0.4)$, as a $9,000\times6,000$ image. 
%
If the total number of tasks $\of{t} \ge 1$ is given as a
parameter,  it is straight forward to split the image into \of{t} sub-images,
each of which is computed in parallel with and independently of the
others. In our case, the image is split into horizontal slices.
Each task is thus four floating-point numbers denoting the region
coordinates, together with two integers denoting the dimensions of the
sub-image to be drawn. The result of the task is a matrix of pixels,
of size $54,000,000 / \of{t}$. 
For instance, using $\of{t}=20$ tasks will
result in 20 sub-images of size $10.3$ Mb each,
assuming each pixel is encoded in four bytes.

@@

% TODO compute ratios for Network
\begin{center}
  \begin{tabular}{|r|r|r|r|r|}
    \hline
    \#cores  &\#tasks & \of{Cores} & \of{Network} \\
    \hline\hline
    2       & 10 & 15.8       (1.86$\times$) &  20.3  (1.45$\times$)      \\
            & 30 & 15.7       (\underline{1.87}$\times$) &  18.7 (1.57$\times$)       \\
            & 100 & 16.1      (1.83$\times$) &  19.8   (1.48$\times$)    \\
            & 1000 & 19.6     (1.50$\times$) &  38.6    (0.76$\times$)  \\
    \hline
    4       & 10 & 9.50       (3.09$\times$)  &  14.4     (2.04$\times$)  \\
            & 30 & 8.26       (\underline{3.56}$\times$)  &  11.4  (2.58$\times$) \\
            & 100 & 8.37      (3.51$\times$)  &  11.4  (2.58$\times$) \\
            & 1000 & 10.6     (2.77$\times$)  &  20.5   (1.43$\times$) \\
    \hline
    8       & 10 & 9.40       (3.13$\times$)  &  12.6    (2.33$\times$)  \\
            & 30 & 4.24       (\underline{6.93}$\times$)  &   7.6  (3.87$\times$)    \\
            & 100 & 4.38      (6.71$\times$)  &   7.5    (3.92$\times$)  \\
            & 1000 & 6.86     (4.29$\times$)  &  11.3    (2.60$\times$)  \\
    \hline
  \end{tabular}
\end{center}
The best timings are achieved for the \of{Cores} configuration, where
communications happen within the same machine and are thus cheaper.
There are two significant differences with respect to the n-queens
benchmark.  On one hand, the number of tasks can be controlled more
easily than in the case of n-queens. We experimentally figured out the
optimal number of tasks to be 30. On the other hand, each computation
result is an image, rather than just an integer as in the case of
n-queens. Consequently, communication costs are much greater. 
In this particular experiment, the total size of the results
transmitted is more than 200 Mb.
